---
title: "UK comparison"
author: "David Henderson"
date: "13/07/2020"
output: 
  html_document:
    theme: journal
    highlight: haddock
    df_print: paged
    code_folding: hide
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12, fig.height = 9,
                      warning = FALSE, message = FALSE)
```

# Introduction

1st attempt to combine data from all 4 UK nations to look at excess deaths in 2020 by week and location of death. The main idea of this file is to show co-authors what we are hoping to get as an output. This may not be possible. As you will see here, the data for E & W is combined and we don't have NI data yet. 

## Packages

R packages required.....

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(curl)
library(readxl)
library(lubridate)
library(forcats)
library(patchwork)
library(ggthemes)

#Helper function
`%nin%` <- negate(`%in%`)

#Short cut for csv output with html tables
my_datatable <- function(x){
  DT::datatable(x, extensions = "Buttons", options = list(dom = "Bfrtip", 
                                                          buttons = c("csv")))
}

#Baseline plot settings
theme_set(theme_minimal(base_family = "Roboto", base_size = 20) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  plot.caption = element_text(colour = "#AAAAAA"),
                  plot.margin = margin(3,15,3,3,"mm")))

#global options for scientific numbers and significant digits.          
options(scipen = 10,
        digits = 1)
```


# England and Wales data

I can't find this data where E & W aren't combined. Big shout to Elizabeth for spotting where we could get the combined numbers. It comes from the "Download this chart" button on Figure 8 of the ONS release [here](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/bulletins/deathsregisteredweeklyinenglandandwalesprovisional/weekending26june2020). 

```{r, message=FALSE, warning=FALSE}
#Assign url and download into a temp file
url <- "https://www.ons.gov.uk/visualisations/dvc891/fig8/datadownload.xlsx"
temp <- tempfile()
temp <- curl_download(url=url, destfile=temp, quiet=FALSE)

#Now read in and tidy the names
ew <- read_xlsx(temp, range = "A8:Q24") %>% 
  rename(home_all = `All deaths...2`,
         home_covid = `COVID-19 deaths...3`,
         home_average = `All deaths 5-year average...4`,
         home_excess = `Excess deaths...5`,
         hospital_all = `All deaths...6`,
         hospital_covid = `COVID-19 deaths...7`,
         hospital_average = `All deaths 5-year average...8`,
         hospital_excess = `Excess deaths...9`,
         carehome_all = `All deaths...10`,
         carehome_covid = `COVID-19 deaths...11`,
         carehome_average = `All deaths 5-year average...12`,
         carehome_excess = `Excess deaths...13`,
         other_all = `All deaths...14`,
         other_covid = `COVID-19 deaths...15`,
         other_average = `All deaths 5-year average...16`,
         other_excess = `Excess deaths...17`)

#Tidy up so that location is a variable
ew %<>% 
  pivot_longer(cols = home_all:other_excess, names_to = "place",
               values_to = "value") %>% 
  separate(place, c("location", "type"), sep = "_")  %>% 
  pivot_wider(names_from = "type", values_from = "value") %>% 
  arrange(`Week no`, location) %>% 
  mutate(location = factor(location,
                           levels = c("hospital", "carehome", "home", "other"),
                           labels = c("Hospital", "Care Home", "Home", "Other")),
         non_covid = all - covid,
         country = "England & Wales") %>% 
  select(location, `Week no`, all, non_covid, everything()) %>%
  rename(week_number = `Week no`) %>%
  mutate(week_number = str_replace(week_number, "Week ", ""),
         week_number = as.integer(week_number)) %>% 
  arrange(location)
ew
```


# Scotland data

This comes from two files on the NRS site. One has weekly COIVD and all-cause deaths, the other the weekly numbers (including averages) for the previous 5 years.

```{r, warning=FALSE, message=FALSE}
temp_1 <- tempfile()
temp_2 <- tempfile()
source <- "https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-data-week-26.zip"

temp_1 <- curl_download(url = source, destfile = temp_1, quiet = FALSE)
unzip(temp_1, exdir = temp_2)

ch_death_covid <- read_csv(file.path(temp_2,"covid-deaths-data-week-26_Table 1 - COVID deaths.csv"), skip = 3)
ch_death_all <- read_csv(file.path(temp_2, "covid-deaths-data-week-26_Table 2 - All deaths.csv"), skip = 3)
#5 year average can be downladed directly from this link...
ch_death_average <- read_csv("https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-2015-2019.csv", skip = 2)

#Now tidy and join together

loc_names <- c("Care Home", "Home and other non institution", "Hospital",
          "Other institution")

# Create a look-up table for week numbers to dates
week_lookup <- tibble(
  week_number = 1:52,
  week_date = seq(ymd(20191230), ymd(20201221), by = "1 week"))


#Wrangle Average deaths
ch_death_average %>% 
  slice(2:6, 9:13, 16:20, 23:27) %>% 
  mutate(location = rep(loc_names, each = 5, times = 1)) %>% 
  select(year = `Week number2`, location, everything(), -`53`) %>% 
  pivot_longer(cols = `1`:`52`, names_to = "week_number",
               values_to = "n_deaths") %>% 
  group_by(location, week_number) %>%
  mutate(min_deaths = min(n_deaths),
         max_deaths = max(n_deaths),
         mean_deaths = mean(n_deaths)) %>% 
  distinct(location, week_number, .keep_all = TRUE) %>% 
  select(-year, -n_deaths) %>% 
  ungroup %>% 
  mutate(week_number = as.integer(week_number)) -> sc

#Wrangle all deaths
ch_death_all %>% 
  select(location = X2, everything(), -`Week beginning`, -X30:-X65) %>% 
  slice(85:88) %>%  
  mutate_at(vars(`30-Dec-19`:`22-Jun-20`), as.numeric) %>% 
  pivot_longer(cols = `30-Dec-19`:`22-Jun-20`, names_to = "date",
               values_to = "deaths_all_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:26, each = 1, times = 4),
         location = rep(loc_names, each = 26, times = 1)) %>% 
  select(-date) %>% 
  left_join(sc, .) -> sc


#Wrangle Covid deaths and join together
ch_death_covid %>% 
  select(location = X2, everything(), -`Week beginning`,
         -`Year to Date`:-X44) %>% 
  slice(83:86) %>%  
  pivot_longer(cols = `30-Dec-19`:`22-Jun-20`, names_to = "date",
               values_to = "deaths_covid_2020") %>% 
  mutate(date = dmy(date),
         week_number = rep(1:26, each = 1, times = 4),
         location = rep(loc_names, each = 26, times = 1)) %>% 
  select(-date) %>% 
  left_join(sc, .) %>% 
  mutate(deaths_nonCovid_2020 = deaths_all_2020 - deaths_covid_2020,
         location = factor(location,
                           levels = c("Hospital", "Care Home", 
                                      "Home and other non institution", 
                                      "Other institution"),
                           labels = c("Hospital", "Care Home", "Home", "Other"))) %>% 
  left_join(., week_lookup) %>% 
  filter(week_number >=11 & week_number <= 26) %>% 
  select(-min_deaths, -max_deaths, -X29, -week_date) %>% 
  rename(all = deaths_all_2020,
         covid = deaths_covid_2020,
         non_covid = deaths_nonCovid_2020,
         average = mean_deaths) %>% 
  mutate(excess = all - average,
         country = "Scotland") -> sc

sc
```


# Join together

```{r}
uk <- full_join(ew, sc)
```

# Plot

This is the sort of plot we want as an end result, but with 4 nations....

```{r}
uk %>% 
  filter(location != "Other") %>% 
  mutate(pct_change = (excess / average * 100)) %>% 
  ggplot(aes(week_number, pct_change, fill = location)) +
  geom_col() +
  facet_grid(country ~ location) +
  scale_fill_wsj() +
  scale_y_continuous(limits = c(-50, 300),
                     labels = scales::percent_format(scale = 1)) +
  theme(legend.position = "none") +
  labs(title = "Percentange change in all-cause mortality 2020",
       subtitle = "from previous 5-year average",
       x = "Week Number",
       y = "",
       caption = "Source: ONS & NRS\nCode:https://github.com/davidhen/ltc_covid_scotland/blob/master/uk_comparator.Rmd")
```

# Data table

Click on the CSV button to download the wrangeled dataset..

```{r}
my_datatable(uk)
```


