---
title: "File 1"
author: "David Henderson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12, fig.height = 9,
                      warning = FALSE, message = FALSE)
```


# Introduction

This file shows the wrangling to raw files taken from: 

  1. The English place
  2. The Scottish one
  3. The Welsh one
  4. You guessed it...
  
These files are used to summarise the number of care homes in each nation and to calculate the total number of beds available. 

## Software

R packages required and helper functions

```{r}
library(tidyverse)
library(curl)
library(readxl)
library(forcats)
library(ggthemes)
library(here)
#Helper function
`%nin%` <- negate(`%in%`)

#Short cut for csv output with html tables
my_datatable <- function(x){
  DT::datatable(x, extensions = "Buttons", options = list(dom = "Bfrtip", 
                                                          buttons = c("csv")))
}
#Set base directory
here()
```


## Read in raw data

These files are stored locally (the English data is quite big) and are available on [the project GitHub pages](https://github.com/davidhen/ltc_covid_uk) in the `raw_data` folder.

```{r}
eng <- read_xlsx(here("raw_data/ENGLAND_care_homes.xlsx"), sheet = "HSCA Active Locations")
scot <- read_xlsx(here("raw_data/SCOTLAND_care_homes.xlsx"), sheet = "MDSF_data_31 May 2020")
ni <- read_xlsx(here("raw_data/NI_care_homes.xlsx"))
w <- read_xlsx(here("raw_data/22 - CIW - 20200730 - list of adult care homes in Wales.xlsx"), 
               range = "A4:K1060")
```

# Wrangle

## England

The ... data contains a lot of information, not just about care homes. For our purposes we will filter using the convenient `Care home?` variable and also to include services that come under the bracket of "Adult social care". We then keep only the required variables and drop the rest before adding a `country` variable to indicate this is English data. 

```{r}
eng %<>% 
  filter(`Care home?` == "Y" & 
           `Location Inspection Directorate` == "Adult social care") %>% 
  select(`Location Name`, `Care homes beds`, `Service user band - Older People`) %>%
  mutate(country = "England")
my_datatable(eng)
```

## Scotland

A similar story with the Scottish Care Inspectorate data. This time we use the "Care Home Service" and keep all service subtypes except "Children & Young People" and drop all inactive services. Again we add the `country` variable.

```{r}
scot %<>% 
  filter(CareService == "Care Home Service" & 
           Subtype %nin% "Children & Young People" &
           ServiceStatus == "Active") %>% 
  select(`Location Name` = ServiceName, 
         `Care homes beds` = TotalBeds, Subtype) %>% 
  mutate(country = "Scotland")
my_datatable(scot)
```

## Northern Ireland

This time keep all Nursing home and Residential home categories and add the `country` variable. 

```{r}
ni %<>% 
  filter(Category %in% c("Nursing (NH)", "Residential (RC)")) %>%
  select(`Location Name` = ServiceName,
         `Care homes beds` = `Max Approved Places`,
         `Categories of Care`) %>% 
  mutate(country = "Northern Ireland")
my_datatable(ni)
```

## Wales

Data here is a little more sparse, we just drop the uneeded variables from the Care Inspectorate Wales file. 

```{r}
w %<>% 
  select(`Location Name` = `Service Name`,
         `Care homes beds` = `Maximum Capacity`,
         `Service Sub Type`) %>% 
  mutate(country = "Wales")
my_datatable(w)
```


# Join

Now we create a `uk` dataframe by binding the previous 4 files together and then coercing the `country` varaible to a factor to make it play better with plots. 

```{r}
uk <- bind_rows(eng, scot, w, ni) %>% 
  mutate(country = factor(country,
                          levels = c("England", "Scotland", "Wales", "Northern Ireland")))
my_datatable(uk)
```

## Save

This object is now written to a `derived_data` folder so we can load it in to another file and start analysing. 

```{r, eval=FALSE}
feather::write_feather(uk, here("derived_data/uk.feather"))
```

# [HOME PAGE](https://davidhen.github.io/ltc_covid_uk/index.html)

```{r}
devtools::session_info()
```


